SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-26 2017-Xudong-SMT/ActaInformatica.tex
prof 2 and th 3

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a pLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State \QIN{$\mathit{L1}$ = \texttt{filterUnsatTransitions}($\mathit{L,Assignements}$)}
\State \QIN{$\mathit{OA}$ = \texttt{makeReachableSubAutomaton}($\mathit{InitState}(P)$,$\mathit{L1}$)}
\State \Return $\mathit{OA}$;

@@

\subsection{Management of State Variable Assignments}

In a PLTS, there may be several incoming transitions 
that assign potentially different (symbolic) values to a state variable.
To handle such cases, the algorithm manages a
list of assignment expressions for each PLTS state. 
\QIN{The only assignments for state variables that happen at the PLTS level are as follows: every time the MakeTransitions
algorithm is applied on a PLTS, it collects the expressions from the $Post$ in each transition.
% As there is no assignments collecting work on a pNet node, it is not
% shown in Algorithm~\ref{alg2}.
}

For a global state in the open automaton, the set of state
variables (which may be used in an outgoing transition) 
is the disjoint union
of the sets of state variables of the individual PLTS states constituting this
global state. \ERIC{The final assignment sets are assembled when building the
open automaton from the reachable transitions.}

@@

@@

\ERIC{The basic algorithm from the previous section potentially explores some
areas of the global state-space that will not be reachable after
checking the OTs' predicates for satisfiability. Incorporating the
satisfiability check inside the residual algorithm will definitely
save some of these useless computation. More precisely, it will still
create, and submit to Z3, the OTs at the ``border'' of the state-space, but
not the transitions issued from non-reachable global states. We use
this idea to create our ``smart'' algorithm.}

But there is more to this. Due to the symbolic nature of the
  transitions, reachability is not as easy to define as in standard
  models. As mentionned above, we collect all
possible (symbolic) assignments of state variables, in the
computed open automaton. By avoiding building the OTs coming from
some unreachable states, we also avoid adding the corresponding assignments
to the variables of their target states. The set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentially more information, which can be used in
the SAT checking of subsequent transitions. But we \emph{cannot} use this
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. So the smart algorithm proceeds in two steps: during the
first step, the SAT checking is done without using any assignment set
information. Then after the completion of this step, we perform another
SAT checking of all transitions from step1, but this time using the
assignment information. This may reduce the reachable state space. If
we want to be sure that the obtained automaton is minimal, we should
iterate this step until a fixed point is reached. But in practical cases, very
little will be gained, so the default implementation does only one
step. Leaving some UNSAT transitions in the
symbolic automaton is still logically correct.

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
%%        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
%%\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\IfThenElse {(NonEmptyAssignmentSet($\mathit{L}$))}
        {\State \quad\ \Return $\mathit{OA}$=makeAutomaton($\mathit{InitState}(P),L)$;}
        {\State \quad\ $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
         \State \quad\ \Return $\mathit{OA}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$};
\end{algorithmic}  
\end{algorithm}

@@

The easy case is when the system does not contain assignments to state
variables. This does not mean ``data-independency'': we can still have
guards in the PLTS transitions and in the synchronisation vectors,
making the behaviour depending on the value of input variables. But
the open automaton needs not to collect the set of possible values dealing
with the values of local variables, so we can directly compute the
reachable part of the automaton using the on-the-fly algorithm.

@@

\begin{proof}
Let $L$, $L1$, $L_{Smart}$ respectively mean the result open transition set of original algorithm before pruning, original algorithm after pruning and smart algorithm. Similarly, $State(L)$, $State(L1)$, $State(L_{Smart})$ mean the result global state set of them. According to the algorithm, there has $L_{Smart}\in L$ and $State(L_{Smart})\in State(L)$.

@@

But in the presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still, the smart strategy will explore a reduced state
  space, not necessarily the minimal one (discussion later).

\begin{theorem}[Correctness with assignments]
  \label{Th:CorrectnessWithAssignments}
In the general case, the smart
algorithm, with a final reachability computation, computes 
the same open automaton as that of Algorithm 1 up to a renaming of  variables in the transition labels.
\end{theorem}

>>>>

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a pLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State \QIN{$\mathit{L1}$ = \texttt{filterUnsatTransitions}($\mathit{L,Assignements}$)}
\State \QIN{$\mathit{OA1}$ = \texttt{makeReachableSubAutomaton}($\mathit{InitState}(P)$,$\mathit{L1}$)}
\State \Return $\mathit{OA1}$;

@@

\subsection{Management of State Variables and Assignments}

In a PLTS, there may be several incoming transitions 
that assign potentially different (symbolic) values to a state variable.
To handle such cases, the algorithm manages a
list of assignment expressions for each PLTS state. 
The only assignments for state variables that happen at the PLTS level
are as follows: every time the MakeTransitions 
algorithm is applied on a PLTS, it collects the expressions from the
$Post$ in each transition. 
% As there is no assignments collecting work on a pNet node, it is not
% shown in Algorithm~\ref{alg2}.

For a global state in the open automaton, the set of state
variables (which may be used in an outgoing transition) 
is the disjoint union
of the sets of state variables of the individual PLTS states constituting this
global state. \ERIC{The final assignment sets are assembled when building the
open automaton from the reachable transitions, see below.}

@@

Function \texttt{filterUnsatTransitions}, in line 11 of
Algorithm\ref{alg1} makes use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far, and
returns the set $L1$ of satisfiable OTs.

\subsection{Building the Open Automaton}

\ERIC{The set $L1$ may contain unreachable transitions, the last step of the
algorithm builds the final Open Automaton as a traversal of $L1$,
starting at the initial state $InitState(p)$ as defined in
page~\pageref{def-states}. More precisely, this is a residual
algorithm building a graph from transitions in $L1$, and states are
the set of target states of transitions in $L1$, plus the initial
states. Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.}

\ERIC{A further reduction may now be gained by using the information from
these sets of possible assignments: adding this information when
checking satisfiability of transitions leaving a given state may
result in more Unsat transitions: the set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentially more information, which can be used in
the SAT checking of subsequent transitions. In turn this may reduce the possible
assignments of the corresponding target states. In theory we should
iterate this procedure, each iteration reducing the set of
transitions, and the sets of assignments; the final Open Automaton
would be the fix point of this iteration. In practice this would be
expensive, and it is not incorrect to leave some Unsat OTs in the
result; in practical cases, very little will be gained computing the
fixpoint, so we only perform one step. Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets.}

@@

The basic algorithm from the previous section potentially explores some
areas of the global state-space that will not be reachable after
checking the OTs' predicates for satisfiability. Incorporating the
satisfiability check inside the residual algorithm will definitely
save some of these useless computation. More precisely, it will still
create, and submit to Z3, the OTs at the ``border'' of the state-space, but
not the transitions issued from non-reachable global states. We use
this idea to create our ``smart'' algorithm.

But there is more to this. Due to the symbolic nature of the
  transitions, reachability is not as easy to define as in standard
  models, as illustrated by the definition above. By avoiding building
  the OTs coming from some unreachable states, we also avoid adding
  the corresponding assignments to the variables of their target states.
But we \emph{cannot} use the
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. We first need to compute a first over-approximation of
the reachable state space, and assignment sets; this corresponds to
the set of OTs $L2$ in Algorithm~\ref{alg3}, line 15.
So the smart algorithm proceeds in two steps: during the
first step, the SAT checking is done without using any assignment set
information. Then after the completion of this step, we perform another
SAT checking of all transitions from step1, using the
assignment information, in the same way it is done at the end of
Algorithm~\ref{alg1}.

There is one case much simpler, that is when the pNets contains no
state variable assignment. Then the set of transitions computed
on-the-fly is already guaranted to be the final reachable set (line 13).

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
%%        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
%%\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\IfThenElse {(NonEmptyAssignmentSet($\mathit{L}$))}
        {\State \quad\ \Return $\mathit{OA2}$=makeAutomaton($\mathit{InitState}(P),L)$;}
        {\State \quad\ $\mathit{L2}$=filterUnsatTransitions($\mathit{L,Assignments}$);
         \State \quad\ \Return $\mathit{OA2}$=makeReachableSubAutomaton($\mathit{InitState}(P),L2)$};
\end{algorithmic}  
\end{algorithm}

@@

The easy case is when the system does not contain assignments to state
variables. This does not mean ``data-independency'': we can still have
guards in the pLTS transitions and in the synchronisation vectors,
making the behaviour depending on the value of input variables. But
the open automaton needs not to collect the set of possible values dealing
with the values of local variables, so we can directly compute the
reachable part of the automaton using the on-the-fly algorithm.

@@

\begin{proof}
  Let $pn=<...>$ be a pNet. The set of states 
  managed by the two algorithms both are subsets of the cartesian
  product of the pLTS states, as defined in \pageref{def-states}. In a
  similar way, the set of state variables are also the same. Comparing
  transitions requires a little more care, because within the
  MakeTransitions function, fresh variables are created everytime a
  synchronisation vector is used (this occurs within the $MkPred$ and
  $matching$ functions). However, given a global state, the call of
  $MakeTransitions$ will always returns the same set of OTs, up to
  renaming of these fresh variables.

  Let us define the following sets of OTs, as computed by the 2
  algorithms:
  \begin{itemize}
    \item $L$ = original set of transitions by Algorithm 1, before pruning
      (i.e. before line 11),
    \item $L1-reach$ = final set of OTs of Algorithm 1, as computed
        by line 12.
      \item $L2-reach$ = final set of OTs of Algorithm 2, as computed
          by line 13.
  \end{itemize}

  For each of these, define $States : OT^* -> State^*$ the function
  computing the set of target states of a set of OTs.

  We want to prove that $L1-reach = L2-reach$ (upto renaming of fresh
  variables) and $States(L1-reach) = States(L2-reach)$.

  Now remark that without assignments,
  Theorem~\ref{Th:CorrectnessWithoutAssignments} boils to proving

Let $L$, $L1$, $L_{Smart}$ respectively mean the result open transition set of original algorithm before pruning, original algorithm after pruning and smart algorithm. Similarly, $State(L)$, $State(L1)$, $State(L_{Smart})$ mean the result global state set of them. According to the algorithm, there has $L_{Smart}\in L$ and $State(L_{Smart})\in State(L)$.

@@

But in the presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still, the smart strategy will explore a reduced state
  space, not necessarily the minimal one (discussion later). In the
  following theorem, both algorithm are considered in their complete
  version, meaning the reachability (with assignments) step is
  iterated up to fixpoint.

\begin{theorem}[Correctness with assignments]
  \label{Th:CorrectnessWithAssignments}
In the general case, the smart algorithm computes 
the same open automaton as that of Algorithm 1 up to a renaming of  variables in the transition labels.
\end{theorem}
